The importance of applying computational creativity\\ to scientific and mathematical domains\\ Position Paper

### 2019-03-05 UnderstandabilityPaper/iccc-science.tex
My own submitted edits

Adding computers to the social process, to form a combination of
people, computers, and mathematical archives to create and apply
mathematics -- a ``mathematics social machine'' \cite{martin:2013} --
further complicates matters. \emph{Automated theorem proving} is the
task of deciding whether a given formal statement follows from a given
set of premises ~\cite{sutcliffe2001evaluating}.  The least
informative approaches, algorithms based on the \emph{resolution}
rule~\cite{robinson1965machine}, produce merely a ``yes'', ``no'' or
``unknown'' response. Not only is this devoid of \emph{explanation},
but it also hides the effects of any bugs; requiring the user to
either trust the results, or verify the implementation.

Both of these issues can be mitigated by having the system instead
generate a \emph{proof object}: a formal argument for \emph{why} a
given statement follows or does not follow. Once generated, a proof
object's validity can be checked without requiring any knowledge of
how it was created, thus avoiding the need to trust or verify the
(potentially complicated) search and generation procedures. The
\emph{de Bruijn criterion}~\cite{barendregt2005challenge} describes
theorem provers which produce proof objects that are trivial to check
by independent \emph{proof checker} programs (which are themselves
easily verified, due to their simplicity); examples are
Coq~\cite{barras1997coq} and Isabelle~\cite{nipkow2002isabelle}.

@@

\subsection{The ``Framing '' approach} 
Enhancing software with explanatory functionality would also help to
mitigate the understandability problem. The ``F'' from the FACE model
stands for {\em Framing}, and we advocate a dual-approach of software
generating framing information alongside an artefact, problem solution
or new data pattern. We foresee this being an increasingly important
area of research in CC, with an increasing level of sophistication:
from explanation to justification to argument and dialogue with a user
about the value, method of production, motivation etc behind
output. How framing information should be developed is a research
programme in its own right. For now, we discuss greater and lesser
understandability in terms of describing the processes underlying the
generative act and consider these for ML approaches.

Many ML approaches can be characterised as constructing a computer
program (or ``model'') consisting of two parts: an overall structure
or \emph{architecture}, which remains fixed; and a set of adjustable
\emph{parameters}, which are inferred or ``learned'' from data
(e.g. \emph{training data} of desired input/output examples). One
particularly simple architecture is the \emph{decision tree}: nested
boolean queries of the input, often used for
\emph{classification}~\cite{safavian1991survey}. These queries are
parameters, and are chosen based on how efficiently they separate the
classes given in the training data.  Decision trees are nominally
understandable, since their behaviour on a given input traces a single
path through these queries; yet they perform poorly compared to other
ML algorithms. The \emph{random forest} approach yields better results
by combining many decision trees and having them vote on the overall
outcome~\cite{randomforests}, although such ensemble behaviour is more
difficult to reason about than that of a single tree.

Recent research has focused on highly expressive classes of models such as
\emph{differentiable programming}~\cite{wang2018demystifying}, whose
architectures output not only a (numerical) answer, but also partial derivatives
with respect to the parameters; and \emph{probabilistic programming}, which
samples from a \emph{distribution} conditioned on the training data. Both
frameworks allow arbitrary architectures, specified via Turing-complete
languages, and provide efficient, composable methods for optimising the
parameters (e.g. Stochastic Gradient Descent and Markov Chain Monte Carlo,
respectively) to minimise arbitrary loss functions (e.g. the squared output
error on a training data set).

With such expressive formalisms, the conflict between the generality
of a model and its understandability becomes clear. Task-specific
architectures require fewer parameters than general-purpose approaches
and perform well with little training data, for example hand-written
characters can be classified based on a single example if we allow our
model to assume the given images are generated by pen
strokes~\cite{lake2015human}. Likewise we can infer parameters of a 3D
scene (such as object position and lighting) from images if we embed a
ray-tracer into our model~\cite{li2018differentiable}; embedding a
physics engine can provide predictions about these scenes, which are
useful e.g. for robot controllers~\cite{degrave2016differentiable}.

The parameters of such programs are easily understood: they describe
the pen strokes, geometric and physical properties of the simulated
system, inferred from the given images. Framing information which
describes the processes underlying the generative act could be
generated; for instance:

\begin{itemize}
\item Decision trees: ``Class {\em x} was chosen because {\em y} was
  greater than {\em z} and ...''
\item Random forests: ``Class {\em x} was chosen because most trees
  looking at {\em y} and {\em z} voted for it, ...''
\item Handwriting recognition: ``{\em B} was chosen because we saw one
  long stroke and two curved strokes''
\item The robot controller: ``The motor was engaged because the
  pendulum had begun to rotate...''
\end{itemize}

However, such specificity makes these implementations
unsuitable for anything else. The choice of such high-level, task-specific
components is performed by the user, and encodes some of their domain knowledge
into the structure of the solution, such that it doesn't have to be learned.
This is similar to how logics like HOL can directly represent high-level
concepts (e.g. natural numbers and sets) but whose proof methods have limited
reusability due to the difficulty of automating such high-level reasoning.

At the other end of the spectrum we have differentiable programs with general
purpose architectures, like \emph{neural networks} (capable of universal
function approximation~\cite{funahashi1989approximate}). These are compositions
of a large number of identical sub-expressions (``neurons''), whose parameters
(``weights'') scale the input values, and hence the contribution of each
sub-expression to the whole. Such architectures encode essentially no domain
knowledge, requiring much larger training sets than task-specific algorithms in
order to ``learn'' these details.  The behaviour of general purpose models
depends so heavily on their (many) parameters, that understanding their
behaviour becomes difficult and they are often treated as inscrutable ``black
boxes''; akin to the large (un)SAT proof described above.

Understanding exactly how these programs make their decisions is an
active area of research, known as \emph{explainable artificial
  intelligence}
(XAI)~\cite{dovsilovic2018explainable,doshi2017towards,molnar2018interpretable},
with popular approaches such as saliency maps~\cite{simonyan2013deep}
tending to reverse-engineer the factors which lead to a model's
decisions (judging the saliency of each pixel based on how strongly it
would effect the output prediction if adjusted). These methods appear
intuitive, e.g. producing visualisations highlighting a particular
object in a scene as the reason for its classification; akin to the
proof objects emitted by theorem provers. Yet this can obscure the
difficulty of interpreting such high-dimensional decision
boundaries. In particular, these explanations or justifications can be
fundamentally altered by imperceptible adjustments to the
input~\cite{ghorbani2017interpretation}. Similar adjustments can also
change a model's output, leading to the field of \emph{adversarial
  machine learning}~\cite{goodfellow2014explaining}; adjustments to
even a single pixel can not only cause a system to mislabel an input,
but to give high confidence to its erroneous
result~\cite{su2019one}. Such fragility and opaqueness has important
implications as these techniques begin to have decision making
applications in life-critical domains such as automotive
control~\cite{bojarski2016end} and medical
diagnosis~\cite{esteva2017dermatologist}.

>>>>

Adding computers to the social process, to form a combination of
people, computers, and mathematical archives to create and apply
mathematics -- a ``mathematics social machine'' \cite{martin:2013} --
further complicates matters. Take \emph{automated theorem proving}, the
task of deciding whether a given formal statement follows from a given
set of premises ~\cite{sutcliffe2001evaluating}.  The least
informative approaches, algorithms based on the \emph{resolution}
rule~\cite{robinson1965machine}, produce merely a ``yes'', ``no'' or
``unknown'' response. Not only is this devoid of \emph{explanation},
but it also hides the effects of any bugs; requiring the user to
either trust the results, or verify the implementation.

This can be mitigated by having the system instead
generate a \emph{proof object}: a formal argument for \emph{why} a
given statement follows or does not follow. Once generated, a proof
object's validity can be checked without requiring any knowledge of
how it was created, thus avoiding the need to trust or verify the
(potentially complicated) search and generation procedures. Theorem provers
which produce proof objects that are trivial to check by independent
\emph{proof checker} programs (which are themselves
easily verified, due to their simplicity) satisfy the
\emph{de Bruijn criterion}~\cite{barendregt2005challenge}; examples are
Coq~\cite{barras1997coq} and Isabelle/HOL~\cite{nipkow2002isabelle}.

@@

\subsection{The ``Framing '' approach}
Enhancing software with explanatory functionality would also help to
mitigate the understandability problem. The ``F'' from the FACE model
stands for {\em Framing}, and we advocate a dual-approach of software
generating framing information alongside an artefact, problem solution
or new data pattern (proof objects can be seen as a limited form of framing
information). We foresee this being an increasingly important area of
research in CC, with an increasing level of sophistication:
from explanation to justification to argument and dialogue with a user
about the value, method of production, motivation etc behind
output. How framing information should be developed is a research programme in
its own right. For now, we discuss greater and lesser understandability in
terms of describing the processes underlying the generative act and consider
these for ML approaches.

Many ML approaches can be characterised as constructing a computer
program (or ``model'') consisting of two parts: an overall structure
or \emph{architecture}, which remains fixed; and a set of adjustable
\emph{parameters}, which are inferred or ``learned'' from data
(e.g. \emph{training data} of desired input/output examples). One
particularly simple architecture is the \emph{decision tree}: nested
boolean queries of the input, often used for
\emph{classification}~\cite{safavian1991survey}. These queries are
parameters, and are chosen based on how efficiently they separate the
classes given in the training data.  Decision trees perform poorly compared to
other ML algorithms, but are nominally understandable since their behaviour on a
given input traces a single path through these queries, which could be turned
into framing information such as ``Class {\em x} was chosen because {\em y} was
greater than {\em z}...''.  The \emph{random forest} approach gives better
performance by combining many decision trees and having them vote on the overall
outcome~\cite{randomforests}, although such ensemble behaviour is more
difficult to reason about than that of a single tree, and is hence harder to
frame in an understandable way. One approach might be to find patterns in the
votes, such as ``Class {\em x} was chosen because most trees looking at features
{\em y} and {\em z} voted for it''.

Recent research has focused on highly expressive classes of models such as
\emph{differentiable programming}~\cite{wang2018demystifying}, whose
architectures output not only a (numerical) answer, but also partial derivatives
with respect to the parameters; and \emph{probabilistic programming}, which
samples from a \emph{distribution} conditioned on the training data. Both
frameworks allow arbitrary architectures, specified via Turing-complete
languages, and provide efficient, composable methods for optimising the
parameters (e.g. Stochastic Gradient Descent and Markov Chain Monte Carlo,
respectively) to minimise arbitrary loss functions (e.g. output error for the
training data).

With such expressive formalisms, the conflict between the generality
of a model and its understandability becomes clear. Task-specific
architectures require fewer parameters than general-purpose approaches, perform
well with little training data, and are amenable to descriptive framing
information. For example, hand-written characters can be classified based on a
single example if we allow our model to assume the given images are generated by
pen strokes~\cite{lake2015human}, and these models may allow descriptions such
as ``Character {\em c} was chosen because there appear to be {\em x} long
strokes, {\em y} curved strokes, etc''. Likewise the parameters of a 3D scene
(such as object position and lighting) can be inferred from images if a
ray-tracer is embedded in the model~\cite{li2018differentiable}; embedding a
physics engine enables predictions about these scenes, which are useful e.g. for
robot controllers~\cite{degrave2016differentiable}. Whilst more complicated than
the previous examples, such a controller could (in principle) justify its
actions based on interpretations of the model, such as ``The motor was engaged
because the pendulum appeared to be falling to the left''.

However, the specificity that makes these implementations understandable also
makes them unsuitable for any other task. The choice of such high-level,
task-specific components is performed by the user, and encodes some of their
domain knowledge into the structure of the solution, such that it doesn't have
to be learned. This is similar to how high-level logics can represent relevant
domain concepts (like natural numbers and sets), yet proof methods making use of
this have limited reusability due to the difficulty of automating such
high-level reasoning.

At the other end of the spectrum are general purpose architectures, like
\emph{neural networks} (differentiable programs capable of universal function
approximation~\cite{funahashi1989approximate}). These are compositions of a
large number of identical sub-expressions (``neurons''), whose parameters
(``weights'') scale their input values, and hence the contribution of each
sub-expression to the whole. Such architectures encode essentially no domain
knowledge, requiring much larger training sets than task-specific algorithms in
order to ``learn'' these details.  So much of these general purpose models'
behaviour comes from tuning their (many) parameters, that understanding or
describing their high-level behaviour is difficult; indeed they are often
treated as inscrutable ``black boxes'', akin to the large (un)SAT proof
described above.

Understanding exactly how such programs make their decisions is an
active area of research, known as \emph{explainable artificial
  intelligence}
(XAI)~\cite{dovsilovic2018explainable,doshi2017towards,molnar2018interpretable}.
Saliency maps are a popular form of framing information~\cite{simonyan2013deep},
which reverse-engineer the factors which lead to a model's decisions (for
example judging the saliency of input pixels by how strongly they each effect
the output prediction if adjusted). These methods appear intuitive, e.g.
producing visualisations highlighting a particular object in a scene as the
reason for its classification; yet this can obscure the difficulty of
interpreting such high-dimensional decision boundaries. In particular,
reasonable justifications (such as classifying an image as a butterfly, with
high saliency given to those pixels which show the butterfly) can be
fundamentally altered by imperceptible adjustments to the
input~\cite{ghorbani2017interpretation} (in this case choosing butterfly based
only on the background vegetation). Similar adjustments can also change a
model's output, leading to the field of \emph{adversarial machine
  learning}~\cite{goodfellow2014explaining}; adjustments to even a single pixel
can not only cause a system to mislabel an input, but to give high confidence to
its erroneous result~\cite{su2019one}. Whilst much attention has been focused on
making the outputs of ML systems more accurate and robust, there is also a need
for framing information which explains more, is more understandable to users and
less prone to misinterpretation.
